---
title: "Summarizing Penguins"
author: "Aida Aquino Arias"
format: gfm
execute: 
  echo: true
---

<!-- If you work with a partner, please include their name with "author" above -->

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(palmerpenguins)
data(penguins)
## loaded the tidyverse package, palmerpenguins package and read in the data
```


## Goals for the Activity  

1. Use `mutate()`, `group_by()`, and `summarize()` to answer some questions about the penguins.   
2. Learn a few other `dplyr` functions such as `drop_na()` and `count()` that can be used in summarizing data.  
3. Learn the `dplyr` functions `case_when()` and `across()` to clean up and create new variables.  
4. Connect summarized and created variables and graphics together to answer research questions.  


**THROUGHOUT THE Activity** be sure to follow the Style Guide by doing the following:  

- load the appropriate packages at the beginning of the Quarto document  
- use proper spacing  
- name all code chunks  
- comment at least once in each code chunk to describe why you made your coding decisions  
- add appropriate labels to all graphic axes  
- use appropriate white space  


## Create new variables with `mutate()`

Use `mutate()` to add a new column, while keeping the existing columns. The general structure is:

```{r}
#| label: mutate-function
#| eval: false
df |>
  mutate(new_column_name = what_it_contains)
```

For example, if I had a data frame df with columns A and B, I can add a new column C that is the sum of A and B as follows (note: you can also use sum(A,B) here instead of A + B):

```{r}
#| label: finding-sum
#| eval: false
df |>
  mutate(C = A + B)
```

For example, we can convert body mass to kilograms:
```{r}
#| label: converting
penguins |>
  mutate(body_mass_kg = body_mass_g / 1000)
## created a new label and ran the given code
```

Notice that for some variables we have `NA` listed for missing values.  We can use `drop_na()` to remove the missing values from those variables

```{r}
#| label: removing-missing-values
penguins |> 
  drop_na() #if we leave it blank it will remove any row with an NA
# created a new label and ran the given code
```


```{r}
#| label: removing-sex
penguins |> 
  drop_na(sex) #adding a variable, removes rows with NA just for that variable
# created a new label and ran the code
```


### Evaluating the Bill Size  

Create new variable called `bill_ratio` that is the ratio of length to depth (length/depth)

```{r}
#| label: creating-new-variable
penguins|>
  mutate(bill_ratio = bill_length_mm/bill_depth_mm)
## created a label and a new varibale called bill_ratio
```


Now create a boxplot of `bill_ratio` split by species.  Be sure to add labels to the axes.  You can add other modifications if you want as well (fill/color, theme, etc).

```{r}
penguins |>
  mutate(bill_ratio = bill_length_mm/bill_depth_mm) |>
  ggplot(aes(x = species,
                   y = bill_ratio,
                   fill = species)) + 
  geom_boxplot() +
  labs(title = "Bill Ratio of Different Species", 
       x = "Species",
       y = "Bill Ratio") + 
  theme_classic() +
  theme_gray()
## created a label and a boxplot that corresponds with the previously mutated code
```


Add three new columns to penguins within one `mutate()` function: one column that contains the bill ratio (bill length / bill depth), one that contains the body mass converted to kg, and one that contains the flipper length converted to meters.

```{r}
#| label: converting-variables
penguins |>
  mutate(bill_ratio = bill_length_mm/bill_depth_mm,
         body_mass_kg = body_mass_g/1000,
         flipper_length_m = flipper_length_mm/1000)
## created new label and filled in the blanks based on the promt
```


We can also use `mutate()` to modify an existing variable.  By default, R reports factors in alphabetical order (you may notice in your boxplot they always print in the same order, Adelie, Chinstrap, Gentoo). If we want to reorder, we can use a function `fct_relevel()` from the `forcats` package, which we will learn about in a few weeks. Modify the code below to calculate the `bill_ratio` and then recreate the boxplot from above by species - how does it differ? 

```{r}
#| label: rearranging-names
library(forcats)
penguins |> 
  mutate(species = 
           fct_relevel(species, "Adelie", "Gentoo", "Chinstrap"),
         bill_ratio = bill_length_mm/bill_depth_mm) |> 
  ggplot(aes(x = species,
                   y = bill_ratio,
                   fill = species)) + 
  geom_boxplot() +
  labs(title = "Bill Ratio of Different Species", 
       x = "Species",
       y = "Bill Ratio") + 
  theme_classic() +
  theme_gray()
## created label and recreated the boxplot from the previous code with the new changes. Some difference are that the colors change based on positions and the names are rearranged.
```


## Summarize data using `group_by()` and `summarize()`

Use the combination of `group_by()` and `summarize()` to find find summary statistics for different groups, and put them in a nice table.

`group_by()` “takes an existing table and converts it into a grouped table where operations are performed ‘by group’”  

`summarize()` “creates a new data frame. It will have one (or more) rows for each combination of grouping variables; if there are no grouping variables, the output will have a single row summarizing all observations in the input. It will contain one column for each grouping variable and one column for each of the summary statistics that you have specified”  

For example, we can calculate the mean and standard deviation for body mass:

```{r}
penguins |>
  group_by(species) |>
  summarize(mass_mean = mean(body_mass_g, na.rm = TRUE),
            mass_sd = sd(body_mass_g, na.rm = TRUE))
```

Note, that `na.rm = TRUE` removes NA from the calculations instead of removing them from the data like `drop_na()`. Often, errors reported with `mean()` and similar functions are because there are `NA` in the data.  

Second Note, if you use `mosaic` in your statistics class, note that `mean()` used here is from base R, so there is no `~` in front of the variable.  If you wanted to use `mosaic` you can, you will just need to modify the syntax:  

- Base R: `mean(x)`  
- Mosaic: `mean(~x)`  



### Group by Species and Sex
Find the mean and standard deviation of body mass in *kilograms* by species and sex.

```{r}
#| label: group-by
penguins |> 
  mutate(body_mass_kg = body_mass_g/1000) |> 
  group_by(species, sex) |> 
  summarize(body_mass_mean = mean(body_mass_kg, na.rm = TRUE),
            bosy_mass_sd = sd(body_mass_kg, na.rm = TRUE))
## created a label and followed the prompt. Using the previous code as a guideline to find the mean and sd of body mass.
```


Starting with penguins, create a summary table containing the maximum and minimum length of flippers (call the columns flip_max and flip_min) for just Chinstrap penguins, grouped by island. (Hint - remember `filter()`)

```{r}
#| label: summary-table
penguins |>
filter(species == "Chinstrap") |>
  group_by(island) |>
  summarize(flip_max = max(flipper_length_mm, na.rm = TRUE), 
            flipper_min = min(flipper_length_mm, na.rm = TRUE))
## created a label, and followed the prompt as close as possible, used the previous code as a guideline
```



Starting with penguins, in a piped sequence:  

- Add a new column called bill_ratio that is the ratio of bill length to bill depth (hint: mutate())  
- Only keep columns species and bill_ratio  
- Group the data by species  
- Create a summary table containing the mean of the bill_ratio variable, by species (name the column in the summary table bill_ratio_mean)  

```{r}
#| label: combining-functions
penguins |>
  mutate(bill_ratio = bill_length_mm/ bill_depth_mm) |>
  select(species, bill_ratio) |>
  group_by(species) |>
  summarise(bill_ratio_mean = mean(bill_ratio, na.rm = TRUE))
## added a new label, and followed the prompt as best as possible, combined multiple functions in order to create a table of specific details
```


## New function `across()`
The `across()` function is especially useful within `summarize()` to efficiently create summary tables with one or more functions applied to multiple variables (columns).

Let’s compare two ways of doing the same thing: creating a summary table of mean values for all penguin size measurements ending in “mm” (bill depth, bill length, flipper length), by species.

Approach 1:
```{r}
penguins |>
  group_by(species) |>
  summarize(bill_length_mean = mean(bill_length_mm, na.rm = TRUE),
            bill_depth_mean = mean(bill_depth_mm, na.rm = TRUE),
            flipper_length_mean = mean(flipper_length_mm, na.rm = TRUE))
```

Approach 2: 
```{r}
penguins |>
  group_by(species) |>
  summarize(across(ends_with("mm"), 
                   .fns = mean, 
                   na.rm = TRUE))
```


We can modify multiple names by appending something to the beginning 
```{r}
penguins |>
  group_by(year) |>
  summarise(across(starts_with("bill"), 
                   .fns = max, 
                   na.rm = TRUE, 
                   .names = "max_{.col}"))
```

or ending of the column name:  
```{r}
penguins |>
  group_by(year) |>
  summarise(across(starts_with("bill"), 
                   .fns = max, 
                   na.rm = TRUE, 
                   .names = "{.col}_max"))
```



Starting from penguins, create a summary table that finds the mean and standard deviation for all variables containing the string “length”, grouped by penguin species. Update the column names to start with “avg_” or “sd_”, followed by the original column names.

There’s quite a bit happening here, so a little breakdown:  

- We use `contains("length")` to indicate we’ll apply the functions to any columns with the word “length” in the name  
- Within `list()` is where the functions to be applied across columns are given, and where their “names” of “avg” and “stdev” are set  
- We use `.names =` to define the final column names in the summary table. Here, the name should start with the function “name” specified above (“avg” or “stdev”), then an underscore, then the original column name (that’s what `"{.fn}_{.col}"` will do)  

```{r}
penguins |>
  group_by(species) |>
  summarize(across(contains("length"),
                   list(avg = mean, stdev = sd),
                   na.rm = TRUE,
                   .names = "{.fn}_{.col}"))
```


### Practice with `across()`
Starting with penguins, group data by island then use `across()` to find the median value of groups for any columns containing the string “mm”. 

Bonus Point: The names in the resulting table should be the original column name followed by an underscore, then the word “median” (e.g. `colname_median`).

```{r}
#| label: across-function
penguins |>
  group_by(island) |>
  summarize(across(contains("mm"), 
                   .fns = median, 
                   na.rm = TRUE,
                   .names = "{.col}_median"))
## added a label and followed the prompt as close as possible. Use the previous code structures as a guideline.
```

Starting from penguins, write a piped sequence to:  

- Exclude penguins observed on Biscoe Island  
- Only keep variables species through `body_mass_g`  
- Rename the species variable to `spp_penguin`  
- Group the data by `spp_penguin`  
- Find the mean value for any variable containing the string “length”, by penguin species, with column names updated to the original column name appended with "_mean" at the end  
```{r}
#| label: excluding-biscoe
penguins |> 
  filter (island != "Biscoe") |>
  select(species : body_mass_g) |>
  rename(spp_penguin = species) |>
  group_by(spp_penguin) |>
  summarize(across(contains("length"),
                   mean, na.rm = TRUE,
                   .names = "{.col}_mean"))
## added new label and followed the prompt while using the past code structures as guidelines.
```


## New Function `count()`

The `dplyr::count()` function wraps a bunch of things into one beautiful friendly line of code to help you find counts of observations by group. To demonstrate what it does, let’s find the counts of penguins in the penguins dataset by species in two different ways:  

- Using `group_by() |> summarize()` with `n()` to count observations  
- Using `count()` to do the exact same thing  

```{r}
penguins |>
  group_by(species) |>
  summarize(n = n())
```

```{r}
penguins |>
  count(species)
```

For example, what does the following code do?
```{r}
penguins |>
  count(species, year)
```

We can combine `count()` with `mutate()` to find proportions and the visualize them - comment the following code:
```{r}
penguins |> 
  count(species, island) |> #count(species, island) |>
  group_by(species) |> #group_by(species)|>
  mutate(total = sum(n),  #mutate(total = sum(n),
         prop = n/total) |>  #prop = n/total)|>
  ggplot(aes(x = island, y = prop, fill = species)) + # ggplot(aes(x = island, y = prop, fill = species)) +
  geom_col(position = position_dodge()) + #geom_col(position = position_dodge()) +
  scale_fill_viridis_d() + #scale_fill_viridis_d()
  labs(x = "Palmer Islands", #labs(x = "Palmer Islands",
       y = "Proportion of Penguin Species per Island", #y = "Proportion of Penguin Species per Island"
       fill = "Species") + #fill = "Species") +
  theme_minimal()#theme_minimal()
## commented the given code, seperately
```

### Practice with `count()`
Starting with penguins, find counts of observation by species, island, and year.
```{r}
#| label: count-species-island-year
penguins |>
  count(species, island, year)
## added a label and found the counts for species, island, and year
```


Starting with penguins, filter to only keep Adelie and Gentoo penguins, then find counts by species and sex.
```{r}
#| label: count-species-and-sex
penguins |>
  filter(species == "Adelie" | species == "Gentoo") |>
  count(species, sex)
## added a label and followed the prompt by filtering Adelie and Gentoo penguins and found the counts of species and sex
```


## New Function `case_when()`
The `case_when()` function is like a really friendly if-else statement. When used within `mutate()`, it allows you to add a new column containing values dependent on your condition(s).

To penguins, add a new column size_bin that contains:  

- “large” if body mass is greater than 4500 g  
- “medium” if body mass is greater than 3000 g, and less than or equal to 4500 g  
- “small” if body mass is less than or equal to 3000 g  

```{r}
penguins |>
  mutate(size_bin = case_when(
      body_mass_g > 4500 ~ "large",
      body_mass_g > 3000 & body_mass_g <= 4500 ~ "medium",
      body_mass_g <= 3000 ~ "small"
    )
  )
```


### Practice with `case_when()`
Identify what is going on in this code: 

```{r}
penguins |>
  select(species, year, flipper_length_mm) |>  # it is only selecting three columns, species, year, flipper_length_mm
  rename(study_year = year) |>   # it is renaming the year column to study_year
  filter(species == "Adelie") |>  #filtering the data to only include the Adelie
  mutate(flipper_rank = case_when(   #mutating the following code
    flipper_length_mm < 200 ~ 1,   #determining if the flipper length is less than 200
    flipper_length_mm >= 200 ~ 2,  #determining if the flipper length is greater than or equal to 200
    TRUE ~ 0 # 0 for anything else
  ))
```


Add a new column to penguins called study_year that contains:  

- “Year 1” if the year is 2007  
- “Year 2” if the year is 2008  
- “Year 3” if the year is 2009 

```{r}
#| label: determine-study-year
penguins |>
  mutate(study_year = case_when(year == 2007 ~ "Year 1", 
                                year == 2008 ~ "Year 2",
                                year == 2009 ~ "Year 3"))
## added a label and assigned the years the corresponding names
```


Starting with penguins, only keep observations for chinstrap penguins, then only keep the `flipper_length_mm` and `body_mass_g` variables. Add a new column called `fm_ratio` that contains the ratio of flipper length to body mass for each penguin. Next, add another column named `ratio_bin` which contains the word “high” if `fm_ratio` is greater than or equal to 0.05, “low” if the ratio is less than 0.05, and “no record” if anything else (e.g. NA). 

```{r}
#| label: flipper-mass-ratio
penguins |>
  filter(species == "Chinstrap") |>
  select(flipper_length_mm, body_mass_g) |>
  mutate (fm_ratio = flipper_length_mm/ body_mass_g, 
          ratio_bin = case_when(!is.na(fm_ratio) & fm_ratio >= 0.05 ~ "high",
           !is.na(fm_ratio) & fm_ratio < 0.05 ~ "low",
           TRUE ~ "no record"))

```


## Wrapping Up
The biggest challenge with data wrangling is figuring out what you want to do first and then thinking about how the vocabulary translates to the code.  Once you know *what* you want to do, it is much easier to look up *how* to do it!

